---
title: "Lab 2 - Intro to ggplot, Using the qplot function"
author: "Jeremy Morris"
output: 
  html_document: 
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
```

# Introduction to ggplot2 - part 1

### About ggplot2
> "ggplot2 is an R package for producing statistical, or data, graphics, but it is unlike most other graphics packages because it has a deep underlying grammar. This grammar, based on the Grammar of Graphics (Wilkinson, 2005), is composed of a set of independent components that can be composed in many different ways. [..] Plots can be built up iteratively and edited later. A carefully chosen set of defaults means that most of the time you can produce a publication-quality graphic in seconds, but if you do have special formatting requirements, a comprehensive theming system makes it easy to do what you want. [..] ggplot2 is designed to work in a layered fashion, starting with a layer showing the raw data then adding layers of annotation and statistical summaries. [..]"

H.Wickham, ggplot2, Use R, DOI 10.1007/978-0-387-98141_1, © Springer Science+Business Media, LLC 2009


# Introduction to ggplot2 - focus on qplot ####

The `ggplot2` library is very flexible and easy to use once you have learned the basics of how it works. There are many ways to put together a visualization using the `ggplot` function which we will cover in the next lab. In this lab, we introduce the `qplot` function which is built to mimic the base R `plot` function. Like the base plot function, there are limits to what the `qplot` function will do. We will focus on creating visualizations with different chart types and adjusting various options to customize those visualizations. At the end of the lab we will show where the `qplot` function starts to produce undesired output.

From the qplot help file:

> qplot is a shortcut designed to be familiar if you're used to base plot(). It's a convenient wrapper for creating a number of different types of plots using a consistent calling scheme. It's great for allowing you to produce plots quickly, but I highly recommend learning ggplot() as it makes it easier to create complex graphics.

We'll work with two datasets:

1. diamonds: data regarding diamond quality
2. mtcars: data regarding cars and fuel efficiency

### diamonds
```{r}
as.tibble(diamonds)
```

### mtcars
```{r mtcars}
as.tibble(mtcars)
```

***

## Creating plots

The main purpose of this discussion is to cover the main options that you will use when creating a plot with `qplot`. We will cover the following topics.

* Basics about the `qplot` function
  * How to specify what goes on the axes
  * How to specify which data will be used
  * How to specify the chart type
* How to change the chart type
* How to add some basic customizations to a chart

### Basic qplot functionality

The following code produces a pretty standard bar chart using the `qplot` function. The various parts of the function call are as follows.

* `clarity` - this is one of the variables from the diamonds data set and functions as the variable of interest.
* `data=diamonds` - the data set to be used when generating the plot.
* `geom=bar` - "geom" stands for "geometric object" and tells `qplot` what kind of chart we're creating (a bar chart in this case).

```{r qplot_hist}
qplot(clarity, data=diamonds, geom="bar")
```

The "bar" geometric object is special in that it nearly always will require some level of aggregation. If you prefer to be specific about the aggregation you are doing, it is best to use the "col" geom as follows. In this example, I have specified each of the options. Note that the "col" geom requires both an x and a y parameter where the "bar" geom only expects an x-axis.

```{r qplot_calc_hist}
diamond_hist <- diamonds %>% group_by(clarity) %>% summarise(n=n())
qplot(x=clarity,y=n,data=diamond_hist,geom='col')
```


### Chart types

Chart types are specified using the `geom` parameter as seen above. When using `qplot`, this just tells the charting engine what to draw. There are many types of geometric objects, we'll cover a subset of the most common.

A complete list of geoms can be found [here](https://ggplot2.tidyverse.org/reference/) in the "Layer: geoms" section and, to a degree, in the "Layer: stats" section.

We will cover the following geoms in this section (noting that bar and col have been already been covered).

* geom_point - scatterplot
* geom_jitter - scatterplot with random noise added
* geom_boxplot - boxplot
* geom_density - density plot
* geom_line - line chart

#### geom_point
A scatterplot can be created by using `geom="point"` or by leaving the option out of the `qplot` call as in the example below.

```{r qplot_scatter}
qplot(wt, mpg, data=mtcars)
```

You can produce the same output by being more specific as in the following example.

```{r qplot_scatter2}
qplot(x=wt, y=mpg, data=mtcars, geom='point')
```

#### geom_jitter
Jitter plots are often used when one of the axes is categorical. First, we'll look at a standard scatterplot.

```{r geom_point_density}
qplot(x=clarity,y=table,data=diamonds,geom='point')
```

This chart does not give us a good view of the full density of points as points layerd on top of each other don't look any different. By changing this to a jitter plot we get a better view of the density of the data points.

```{r geom_jitter}
qplot(x=clarity,y=table,data=diamonds,geom='jitter')
```

#### geom_boxplot
Box plots may not be too common in business presentations, however, I like them a lot so I will cover them quickly here. The boxplot draws a box around the middle half of the data. The bottom of the box starts at the 25th percentile and the top at the 75th with a line in the middle for the 50th (or median). A line is used for data points within the inter-quartile range (IQR) and individual dots are used outside 1.5*IQR. I think they are an excellent way to visualize the density of data points. You can see how it makes the chart a lot cleaner than in the case of the jitter plot.

```{r geom_boxplot}
qplot(x=clarity,y=table,data=diamonds,geom='boxplot')
```


#### geom_density
Also not terribly common in business presentations but very useful for understanding your data. The density only requires an x parameter and requires you to use a fill or facet to visualize multiple dimensions (more on those in future labs). The following shows a single density plot.

```{r geom_density_single}
qplot(x=depth,data=diamonds,geom='density')
```

#### geom_line
Standard line chart. This chart requires an x and a y parameter but also allows you to perform aggregations if you really want to. I prefer to pre-aggregate data as it is often much clearer to someone reading through the code. By someone, I mean my future self who is in some ways smarter than my current self. Line charts are most often used with a date value in the x-axis. Otherwise, there should be some reason that you are connecting the lines together. Here, we'll use the NYC flight data since it includes a date component.

```{r geom_line}
flights_by_day <- nycflights13::flights %>% 
  mutate(dt=as.Date(paste(year,month,day,sep='-'),format='%Y-%m-%d')) %>%
  group_by(dt) %>% summarise(n=n())

qplot(x=dt,y=n,data=flights_by_day,geom='line')
```

As a little bonus, you'll notice that the data is highly seasonal within the week. You can apply various transforms to get the data to a different time grain as you'll see in the following example. Here, I use the `cut` function to truncate or round the dates to the nearest week.

```{r geom_line_week}
flights_by_week <- flights_by_day %>% mutate(dt_week=as.Date(cut(dt,breaks='week'))) %>%
  group_by(dt_week) %>% summarise(n=sum(n))

qplot(x=dt_week,y=n,data=flights_by_week,geom='line')
```


### Adding customizations to the chart

Add aesthetic mapping (i.e. color), recall that `qsec` is an attribute in the mtcars dataset indicating the car's quarter-mile time. Also note that `qplot` recognizes that `qsec` is a quantitative variable and uses a continuous color scale. In the second example, we've used the `as.factor` function to force `qplot` to make `cyl` categorical.
```{r qplot_aes}
qplot(wt, mpg, data=mtcars, color=qsec)
qplot(wt, mpg, data=mtcars, color=as.factor(cyl))
```

Change size of points. In the following example, we use a static size for the points.
```{r qplot_size_static}
qplot(wt, mpg, data=mtcars, color=qsec, size=5)
```

In this example, we vary the size off the points based off of the number of cylinders.
```{r qplot_size_var}
qplot(wt, mpg, data=mtcars, color=qsec, size=cyl) 
```

Use alpha blending changes the transparency of each of the points. This can be useful in some cases, especially if you have a large number of points.
```{r qplot_alpha}
qplot(wt, mpg, data=mtcars, alpha=qsec, size=5)
```

The following example show how to use the following aesthetic mappings.

* `shape` - controls the shape of each point.
* `size` - controls the size of each shape. This can be static or variable if given a column of data.
* `color` - colors in the points. Note that this is different than `fill` which controls the interior color of an object.

```{r qplot_aes2}
qplot(wt, mpg, data=mtcars, shape=factor(cyl), size=5)
qplot(wt, mpg, data=mtcars, size=qsec)
qplot(wt, mpg, data=mtcars, size=qsec, color=factor(carb))
qplot(wt, mpg, data=mtcars, size=qsec, shape=factor(cyl))
qplot(wt, mpg, data=mtcars, size=qsec, shape=factor(cyl), color=factor(cyl)) 
```

Bar-plot
```{r qplot_bar}
qplot(factor(cyl), data=mtcars)
```

Flip plot by 90°. Here we actually blend the use of `qplot` and a `ggplot` function.
```{r qplot_bar_flip}
qplot(factor(cyl), data=mtcars) + coord_flip()
```

The following shows the difference between fill/color bars. Note that `fill` shades the interior of the object while `colour` modifies the line around the object. Also, the author of the `ggplot2` library is from New Zealand, hence the odd spelling of the word 'colour'. The more americanized version 'color' is also available.
```{r qplot_bar_color}
qplot(factor(cyl), data=mtcars, geom="bar", fill=factor(cyl))
qplot(factor(cyl), data=mtcars, geom="bar", colour=factor(cyl))
```

The following shows that if you set `fill` to a different variable, you will get a stacked bar chart. Grouped bar charts can also be created if you're into that sort of thing.
```{r qplot_bar_fill}
qplot(factor(cyl), data=mtcars, geom="bar", fill=factor(gear))
```


The histogram geom allows you to change the binwidths for more precision.
```{r qplot_hist_bins}
qplot(carat, data=diamonds, geom="histogram", binwidth=0.1)
qplot(carat, data=diamonds, geom="histogram", binwidth=0.01)
```

An array of geoms can be passed to the geom parameter creating more sophisticated plots.
```{r qplot_combine}
qplot(wt, mpg, data=mtcars, geom="point")
qplot(wt, mpg, data=mtcars, geom="smooth")
qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"))
```

Tweaking the smooth plot ("loess"-method: polynomial surface using local fitting)
```{r qplot_smooth}
qplot(wt, mpg, data=mtcars, geom="smooth")
```

Removing standard error
```{r qplot_smooth_se}
qplot(wt, mpg, data=mtcars, geom="smooth", se=FALSE)
```

Making line more or less wiggly (span: 0-1)
```{r qplot_smooth_span}
qplot(wt, mpg, data=mtcars, geom="smooth", span=0.6)
qplot(wt, mpg, data=mtcars, geom="smooth", span=1)
```

Using linear modelling
```{r qplot_lm}
qplot(wt, mpg, data=mtcars, geom="smooth", method="lm")
```

Sometimes qplot doesn't give us what we're looking for
```{r qplot_fail}
qplot(wt, mpg, data=mtcars, color=factor(cyl), geom=c("smooth", "point"))
```

Need to upgrade to `ggplot`
```{r ggplot}
ggplot() +
  geom_point(data=mtcars, aes(x=wt, y=mpg, colour=factor(cyl))) +
  geom_smooth(data=mtcars, aes(x=wt, y=mpg))
```

`ggplot2` official documentation
HTTP://docs.ggplot2.org/current/