---
title: "Lab 2 - Intro to ggplot, Using the qplot function"
author: "Jeremy Morris"
date: "2/12/2019"
output: 
  html_document: 
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Introduction to ggplot2 - part 1

### About ggplot2
> "ggplot2 is an R package for producing statistical, or data, graphics, but it is unlike most other graphics packages because it has a deep underlying grammar. This grammar, based on the Grammar of Graphics (Wilkinson, 2005), is composed of a set of independent components that can be composed in many different ways. [..] Plots can be built up iteratively and edited later. A carefully chosen set of defaults means that most of the time you can produce a publication-quality graphic in seconds, but if you do have special formatting requirements, a comprehensive theming system makes it easy to do what you want. [..] ggplot2 is designed to work in a layered fashion, starting with a layer showing the raw data then adding layers of annotation and statistical summaries. [..]"

H.Wickham, ggplot2, Use R, DOI 10.1007/978-0-387-98141_1, © Springer Science+Business Media, LLC 2009


# Introduction to ggplot2 - focus on qplot ####

The `ggplot2` library is very flexible and easy to use once you have learned the basics of how it works. There are many ways to put together a visualization using the `ggplot` function which we will cover in the next lab. In this lab, we introduce the `qplot` function which is built to mimic the base R `plot` function. Like the base plot function, there are limits to what the `qplot` function will do. We will focus on creating visualizations with different chart types and adjusting various options to customize those visualizations. At the end of the lab we will show where the `qplot` function starts to produce undesired output.

From the qplot help file:

> qplot is a shortcut designed to be familiar if you're used to base plot(). It's a convenient wrapper for creating a number of different types of plots using a consistent calling scheme. It's great for allowing you to produce plots quickly, but I highly recommend learning ggplot() as it makes it easier to create complex graphics.

We'll work with two datasets:

1. diamonds: data regarding diamond quality
2. mtcars: data regarding cars and fuel efficiency

### diamonds
```{r}
as.tibble(diamonds)
```

### mtcars
```{r mtcars}
as.tibble(mtcars)
```

***

## Creating plots

The main purpose of this discussion is to cover the main options that you will use when creating a plot with `qplot`. We will cover the following topics.

* How to choose a chart type
* How to specify what goes on the axes
* How to specify which data will be used
* How to add basic options to the chart

### Chart types

Chart types are specified using the `geom` parameter. We will see this in more detail when we get to the more advanced `ggplot` call, but a `geom` is short for a geometric object. When using `qplot`, this just tells the charting engine what to draw. There are two types

The following code produces a pretty standard histogram using the `qplot` function. The various parts of the function call are as follows.

* `clarity` - this is one of the variables from the diamonds data set.
* `data=diamonds` - here we tell the function which dataframe we will use to generate the plot.
* `fill=cut` - the fill parameter adds an additional series to the plot that will be "filled in" with colors depending on the values in the variable specified (`cut` in this case).
* `geom=bar` - "geom" stands for "geometric object" and tells `qplot` what kind of chart we're creating (a bar chart in this case).

One thing not immediately obvious about this call is what kind of aggregation will be done on the data. When no aggregation is specified for a bar chart, all `ggplot2` functions will perform a count as the aggregation method. That's how the following code generates a histogram.

```{r qplot_hist}
qplot(clarity, data=diamonds, fill=cut, geom="bar")
```

If you wanted to do the aggregation explicitely, you could do the following (still using `qplot` for the chart). In this case, we use the `weight` parameter to specify the y-axis. Note, this produces the same output as the prior example.

```{r qplot_calc_hist}
diamond_hist <- diamonds %>% group_by(clarity,cut) %>% summarise(n=n())
qplot(x=clarity,y=n,fill=cut,data=diamond_hist,geom='col')
```

The same chart can be generated using `ggplot` as you can see below. We will cover in more detail how this is done in next week's lab.
```{r ggplot_hist}
ggplot(diamonds, aes(clarity, fill=cut)) + 
    geom_bar()
```

## How to Use `qplot`

To create a scatterplot, you do not need to specify a "geom" in the call. "geom" is set to 'point' by default. The following call sets `x=wt` and `y=mpg`.

Note: By default, `qplot` will generate a bar plot if only one measure is specified and a scatterplot if two are specified. The `geom` parameter must be used if another chart type is desired.

```{r qplot_scatter}
qplot(wt, mpg, data=mtcars)
```

`qplot` allows you to transform input data with functions. Note that the transformations are specified in the axis titles.
```{r qplot_transform}
qplot(log(wt), mpg - 10, data=mtcars)
```

Add aesthetic mapping (i.e. color), recall that `qsec` is an attribute in the mtcars dataset indicating the car's quarter-mile time. Also note that `qplot` recognizes that `qsec` is a quantitative variable and uses a continuous color scale. In the second example, we've used the `as.factor` function to force `qplot` to make `cyl` categorical.
```{r qplot_aes}
qplot(wt, mpg, data=mtcars, color=qsec)
qplot(wt, mpg, data=mtcars, color=as.factor(cyl))
```

Change size of points. In the following example, we use a static size for the points.
```{r qplot_size_static}
qplot(wt, mpg, data=mtcars, color=qsec, size=5)
```

In this example, we vary the size off the points based off of the number of cylinders.
```{r qplot_size_var}
qplot(wt, mpg, data=mtcars, color=qsec, size=cyl) 
```

Use alpha blending changes the transparency of each of the points. This can be useful in some cases, especially if you have a large number of points.
```{r qplot_alpha}
qplot(wt, mpg, data=mtcars, alpha=qsec, size=5)
```

The following example show how to use the following aesthetic mappings.

* `shape` - controls the shape of each point.
* `size` - controls the size of each shape. This can be static or variable if given a column of data.
* `color` - colors in the points. Note that this is different than `fill` which controls the interior color of an object.

```{r qplot_aes2}
qplot(wt, mpg, data=mtcars, shape=factor(cyl), size=5)
qplot(wt, mpg, data=mtcars, size=qsec)
qplot(wt, mpg, data=mtcars, size=qsec, color=factor(carb))
qplot(wt, mpg, data=mtcars, size=qsec, shape=factor(cyl))
qplot(wt, mpg, data=mtcars, size=qsec, shape=factor(cyl), color=factor(cyl)) 
```

Bar-plot
```{r qplot_bar}
qplot(factor(cyl), data=mtcars)
```

Flip plot by 90°. Here we actually blend the use of `qplot` and a `ggplot` function.
```{r qplot_bar_flip}
qplot(factor(cyl), data=mtcars) + coord_flip()
```

The following shows the difference between fill/color bars. Note that `fill` shades the interior of the object while `colour` modifies the line around the object. Also, the author of the `ggplot2` library is from New Zealand, hence the odd spelling of the word 'colour'. The more americanized version 'color' is also available.
```{r qplot_bar_color}
qplot(factor(cyl), data=mtcars, geom="bar", fill=factor(cyl))
qplot(factor(cyl), data=mtcars, geom="bar", colour=factor(cyl))
```

The following shows that if you set `fill` to a different variable, you will get a stacked bar chart. Grouped bar charts can also be created if you're into that sort of thing.
```{r qplot_bar_fill}
qplot(factor(cyl), data=mtcars, geom="bar", fill=factor(gear))
```


The histogram geom allows you to change the binwidths for more precision.
```{r qplot_hist_bins}
qplot(carat, data=diamonds, geom="histogram", binwidth=0.1)
qplot(carat, data=diamonds, geom="histogram", binwidth=0.01)
```

An array of geoms can be passed to the geom parameter creating more sophisticated plots.
```{r qplot_combine}
qplot(wt, mpg, data=mtcars, geom="point")
qplot(wt, mpg, data=mtcars, geom="smooth")
qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"))
```

Tweaking the smooth plot ("loess"-method: polynomial surface using local fitting)
```{r qplot_smooth}
qplot(wt, mpg, data=mtcars, geom="smooth")
```

Removing standard error
```{r qplot_smooth_se}
qplot(wt, mpg, data=mtcars, geom="smooth", se=FALSE)
```

Making line more or less wiggly (span: 0-1)
```{r qplot_smooth_span}
qplot(wt, mpg, data=mtcars, geom="smooth", span=0.6)
qplot(wt, mpg, data=mtcars, geom="smooth", span=1)
```

Using linear modelling
```{r qplot_lm}
qplot(wt, mpg, data=mtcars, geom="smooth", method="lm")
```

Sometimes qplot doesn't give us what we're looking for
```{r qplot_fail}
qplot(wt, mpg, data=mtcars, color=factor(cyl), geom=c("smooth", "point"))
```

Need to upgrade to `ggplot`
```{r ggplot}
ggplot() +
  geom_point(data=mtcars, aes(x=wt, y=mpg, colour=factor(cyl))) +
  geom_smooth(data=mtcars, aes(x=wt, y=mpg))
```

`ggplot2` official documentation
HTTP://docs.ggplot2.org/current/